---
title: "Motor Trend MPG Data Analysis"
author: "Thimmaraju Rudrappa"
date: "August 9, 2019"
output: html_document
---
*Created with knitr*

#### Executive Summary
The objective of this report is to find out which transmission type is fuel efficient interms of Miles per gallon (MPG) by aunderstanding the relationship between transmission type (manual or 
automatic) and MPG using 'mtcars' data set in R. t-test for transmission type (automatic and manual) showed that manual transmission gives 7.245 more MPG than automatic 
transmission vehicles. However multiple linear regressions analysis 
showed that the manual transmission contributed less significantly to MPG and the advantage is only
an improvement of 1.81 MPG.  Other variables such as Weight, horsepower, and number of 
cylinders seems to be contributing more significantly to overall MPG of vehicles.

#### Loading the Data
Load the dataset and convert categorical variables to factors.
```{r results='hide', message=FALSE}
library(ggplot2)
data(mtcars)
head(mtcars, n=3)
dim(mtcars)
mtcars$cyl <- as.factor(mtcars$cyl)
mtcars$vs <- as.factor(mtcars$vs)
mtcars$am <- factor(mtcars$am)
mtcars$gear <- factor(mtcars$gear)
mtcars$carb <- factor(mtcars$carb)
attach(mtcars)
```

#### Exploratory dat Analysis
**See Appendix Figure I** Box plot for exploratory analysis that compares MPG from Automatic Vs Manual 
transmission vehicles. By looking at the plot one can conclude that there is a significant 
increase in MPG for vehicles with a manual transmission compared to automatic transmission vehicles.

##### Statistical analysis 
t-test for transmission types
```{r}
testResults <- t.test(mpg ~ am)
testResults$p.value
```
Null hypothesis is rejected as the difference between transmission types is more than 0.  
```{r}
testResults$estimate
```
The difference estimate between the 2 transmissions is 7.24494 MPG in favor of 
manual.

##### Regression Analysis
Fit the full model of the data
```{r results='hide'}
fullModelFit <- lm(mpg ~ ., data = mtcars)
summary(fullModelFit)  # results hidden
summary(fullModelFit)$coeff  # results hidden
```
Since most of the the coefficients have a p-value more than 0.05, hence we may not able to conclude
which variables are statistically significant. 

selection to identify which variables are statistically significant
```{r results='hide'}
stepFit <- step(fullModelFit)
summary(stepFit) # results hidden
summary(stepFit)$coeff # results hidden
```

New model with 4 variables (cylinders, horsepower, weight, transmission) has the R-squared value of 0.8659 hence it can be concluded that this explains about 87% of the variance in MPG. The coefficients are statistically significant as they have a p-value less than 0.05. From this analysis it can be concluded that increasing the number of cylinders (4 to 6) decreases the MPG by 3.03.  Further increase of cylinders to 8 will decrease the MPG by 2.16. Every 100 horse power increase is expected to decrease the MPG  by 3.21. Each 1000 lbs increase in weight appears to cause a decrease of MPG by 2.5. However Manual transmission appears to improve the MPG by 1.81.

#### Residuals & Diagnostics
Residual Plots
**See Appendix Figure II**

The residual plots conclude the following:

1. Randomness in the Residuals vs. Fitted plot supports the assumption of independence
2. Points on the Q-Q plot aligning to the line shows that the distribution of residuals is normal
3. Random distribution of the Scale-Location plot confirms the constant variance assumption
4. All the points of the Residuals vs Leverage are with 0.05 lines, concludes that there are no outliers
```{r}
sum((abs(dfbetas(stepFit)))>1)
```

#### Conclusion
Manual transmission vehicles have slightly higher MPG compared automatic transmission. However the weight, horsepower & number of cylinders play a critical role in determining fuel efficiency in terms of MPG.

### Appendix Figures

#### Figure 1
```{r echo=FALSE}
  boxplot(mpg ~ am, 
          xlab="Transmission Type (0 = Automatic, 1 = Manual)", 
          ylab="MPG",
          main="MPG by Transmission Type")
```

#### Figure 2
```{r echo=FALSE}
par(mfrow = c(2, 2))
plot(stepFit)
```